---
title: "Making sense of non-parametric tests"
date: 2026-08-27
draft: true
categories:
  - r
  - statistics
execute:
  eval: true
---

When I was studying physics we had our required share of lab work, which was
also our introduction to experimental statistics. It was there that I heard
for the first time about hypothesis testing, in particular about the t-test.
And since I never formally studied Statistics, this is what I always
assumed to be one of the most fundamental tests there was. How wrong.

The problem with the t-test, which is often overlooked, lies in the assumptions
it relies on---particularly the normality assumption. Violations of this
assumption, we're told, can introduce bias. This makes it sound as if the
error is minor at worst; in fact it can lead to completely wrong conclusions.

Imagine we have an experiment with 300 pairs of matched subjects, where one
member of each pair, chosen at random, gets some treatment. For each pair $i$
we measure $d_i = r_{Ti} - r_{Ci}$, the difference in responses between the
treated unit and the control unit. Let's assume now that $d_i$ isn't
normally distributed but, say, lognormally distributed up to some constant:

```{r}
set.seed(20260826)

n <- 300
d <- rlnorm(n, meanlog = 0, sdlog = 1) - 1.4

w <- wilcox.test(
  d,
  mu = 0,
  conf.int = TRUE,
  exact = FALSE
)

c(
  mean   = mean(d),
  median = median(d),
  HL     = unname(w$estimate)
)
```
```{r}
#| echo: false
library(ggplot2)

reference_lines <- data.frame(
  statistic = factor(
    c("Mean", "Median", "Hodges–Lehmann"),
    levels = c("Mean", "Median", "Hodges–Lehmann")
  ),
  value = c(
    mean(d),
    median(d),
    unname(w$estimate)
  )
)

ggplot(data.frame(d), aes(x = d)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 35,
    fill = "grey85",
    colour = "white"
  ) +
  geom_density(
    linewidth = 0.8,
    colour = "grey25"
  ) +
  geom_vline(
    data = reference_lines,
    aes(
      xintercept = value,
      colour = statistic,
      linetype = statistic
    ),
    linewidth = 0.9
  ) +
  scale_colour_manual(values = c(
    "Mean" = "#D55E00",
    "Median" = "#0072B2",
    "Hodges–Lehmann" = "#009E73"
  )) +
  scale_linetype_manual(values = c(
    "Mean" = "solid",
    "Median" = "dashed",
    "Hodges–Lehmann" = "dotdash"
  )) +
  labs(
    x = expression(d[i]),
    y = "Density",
    colour = NULL,
    linetype = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

As you'd expect from a log-normal distribution, there's a heavy tail and the
mean is substantially larger than the median. In fact they are of opposite
signs!

What is a point estimate anyway? Statistics 101 textbooks define it as the value
that maximizes the likelihood. The likelihood of what? Of the model that you
*assume* describes the distribution of the data. An equivalent definition is
that it's an unbiased estimator of the parameter you're trying to estimate.
What parameter? The parameter of the model that you *assume* describes the
distribution of the data. See where I'm going with this?

In the plot above we show three reference lines:

- the median: gives you the minimum effect size for half the population
- the mean: gives you the average effect size
- the Hodges-Lehmann estimator

The HL estimator is, strictly speaking, only useful when the distribution is
symmetric. Like the mean or the median it is a location estimator. It is
robust against outliers (unlike the mean) but more reliable than the median.
Use it for "serious" decisions like launch decisions, but stick to the median
in stakeholder presentations.

It bears saying it again: which estimator to use depends on the application.
You want to use the mean if you worry about the average impact of your
treatment; you want to use the median or the HL estimator if you worry about
the impact to the middle customer experience. This is a product decision, not
a data science one.

One a final note, remember that you can always use bootstrapping to estimate
confidence intervals of whatever estimator you choose to use:

```{r}
hl <- function(x) {
  walsh <- outer(x, x, "+") / 2
  median(walsh[upper.tri(walsh, diag = TRUE)])
}

estimate_location <- function(x) {
  c(
    Mean   = mean(x),
    Median = median(x),
    HL     = hl(x)
  )
}

set.seed(20260828)

B <- 5000

boot_estimates <- replicate(B, {
  d_boot <- sample(d, size = length(d), replace = TRUE)
  estimate_location(d_boot)
})

# Percentile bootstrap intervals
bounds <- t(apply(
  boot_estimates,
  1,
  quantile,
  probs = c(0.025, 0.975),
  names = FALSE
))

observed <- estimate_location(d)

results <- data.frame(
  estimand = names(observed),
  estimate = unname(observed),
  lower_95 = bounds[, 1],
  upper_95 = bounds[, 2],
  row.names = NULL
)

results
```

This shows that the CIs for the median and HL estimator are the tightest, almost
half the width of the mean's CI:

```{r}
#| echo: false
library(ggplot2)

plot_data <- transform(
  results,
  estimand = factor(
    estimand,
    levels = c("HL", "Median", "Mean")
  )
)

ggplot(
  plot_data,
  aes(
    x = estimate,
    y = estimand,
    colour = estimand
  )
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    colour = "grey50"
  ) +
  geom_segment(
    aes(
      x = lower_95,
      xend = upper_95,
      yend = estimand
    ),
    linewidth = 1.2
  ) +
  geom_point(size = 3.5) +
  scale_colour_manual(values = c(
    "Mean"   = "#D55E00",
    "Median" = "#0072B2",
    "HL"     = "#009E73"
  )) +
  labs(
    title = "Three definitions of the treatment effect",
    subtitle = "Point estimates with 95% percentile-bootstrap confidence intervals",
    x = expression("Estimated effect (" * d * ")"),
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )
```

