---
title: "How many rules does it take to guess a German noun's gender?"
date: 2026-09-04
categories:
  - r
  - machine learning
  - german
execute:
  warning: false
  message: false
---

German noun gender is a nice case where the most accurate model is not necessarily
the most useful one. The object I want at the end of this analysis is a single
decision tree small enough for a learner to remember.

That gives the exercise two objectives:

1. predict the right article for nouns one is likely to encounter in practice;
2. use as few intelligible rules as possible.

The output will therefore be a Pareto frontier of predictive accuracy against tree
size, followed by one deliberately small tree. A flexible model reported by
[Fedden et al.](https://openresearch.surrey.ac.uk/esploro/outputs/journalArticle/Typology-meets-statistical-modeling-The-German/991154296202346)
reached about 96% accuracy with much richer phonological, semantic, derivational,
and inflectional information. I will use that number as a research-grade reference,
not as a theoretical upper bound or a directly comparable score.

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(knitr)
library(purrr)
library(readr)
library(rpart)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

theme_set(theme_minimal(base_size = 13))
```

## Download the dictionary and subtitle frequencies

The labels come from the
[german-nouns](https://github.com/gambolputty/german-nouns) export of German
Wiktionary. Its CSV contains roughly 100,000 entries, including gender and
inflected forms, and is published under CC BY-SA 4.0.

The usage weights come from Matthias Buchmeier's
[German subtitle frequency list](https://en.wiktionary.org/wiki/User:Matthias_Buchmeier#German_frequency_list).
It was made from 25,399,099 words of movie and television subtitles and preserves
letter case. The list is split into pages of 5,000 entries; the first ten pages are
enough for this experiment.

Two of the proposed rules depend on recognizing verbs. For those I use the
infinitive column from the Wiktionary-derived
[german-verbs-database](https://github.com/viorelsfetea/german-verbs-database).
It lets me distinguish a spelling test from a semantic claim: a noun can look
exactly like an infinitive or verb stem without actually being derived from that
verb.

The following chunk downloads about 21 MB in total and caches it in this post's
`data` directory. That directory is ignored by Git.

```{r}
#| label: download-data

data_dir <- "data"
dir.create(data_dir, showWarnings = FALSE)

noun_url <- paste0(
  "https://raw.githubusercontent.com/gambolputty/german-nouns/",
  "main/german_nouns/nouns.csv"
)
noun_path <- file.path(data_dir, "nouns.csv")

if (!file.exists(noun_path)) {
  download.file(noun_url, noun_path, mode = "wb", quiet = TRUE)
}

verb_url <- paste0(
  "https://raw.githubusercontent.com/viorelsfetea/",
  "german-verbs-database/master/output/verbs.csv"
)
verb_path <- file.path(data_dir, "verbs.csv")

if (!file.exists(verb_path)) {
  download.file(verb_url, verb_path, mode = "wb", quiet = TRUE)
}

starts <- seq(1, 45001, by = 5000)
ends <- starts + 4999
page_names <- sprintf("German_frequency_list-%d-%d", starts, ends)
subtitle_urls <- sprintf(
  paste0(
    "https://en.wiktionary.org/w/index.php?title=",
    "User:Matthias_Buchmeier/%s&action=raw"
  ),
  page_names
)
subtitle_paths <- file.path(data_dir, paste0(page_names, ".txt"))

walk2(subtitle_urls, subtitle_paths, function(url, path) {
  if (!file.exists(path)) {
    download.file(URLencode(url), path, mode = "wb", quiet = TRUE)
  }
})
```

For a fresh R installation, these are the required packages:

```r
install.packages(c(
  "dplyr", "ggplot2", "knitr", "purrr", "readr", "rpart",
  "scales", "stringr", "tibble", "tidyr"
))
```

## Build a frequency-weighted noun dictionary

The dictionary includes affixes, proper names, phrases, and nouns with more than
one possible gender. I keep ordinary one-word nouns whose Wiktionary record has
exactly one gender.

```{r}
#| label: clean-dictionary

gender_columns <- c("genus", paste("genus", 1:4))

nouns_raw <- read_csv(
  noun_path,
  col_types = cols(.default = col_character()),
  show_col_types = FALSE
)

nouns <- nouns_raw |>
  filter(
    pos == "Substantiv",
    str_detect(lemma, "^[A-ZÄÖÜ][[:alpha:]]+$")
  ) |>
  select(lemma, all_of(gender_columns)) |>
  pivot_longer(
    cols = all_of(gender_columns),
    names_to = NULL,
    values_to = "gender"
  ) |>
  filter(gender %in% c("m", "f", "n")) |>
  distinct(lemma, gender) |>
  add_count(lemma, name = "number_of_genders") |>
  filter(number_of_genders == 1) |>
  transmute(
    lemma,
    gender = factor(
      recode(gender, m = "der", f = "die", n = "das"),
      levels = c("der", "die", "das")
    )
  )

inflection_columns <- names(nouns_raw) |>
  keep(
    ~ str_detect(
      .x,
      "^(nominativ|genitiv|dativ|akkusativ) (singular|plural)"
    )
  )

noun_forms <- nouns_raw |>
  filter(pos == "Substantiv", lemma %in% nouns$lemma) |>
  select(lemma, all_of(inflection_columns)) |>
  pivot_longer(
    -lemma,
    names_to = "inflection",
    values_to = "form",
    values_drop_na = TRUE
  ) |>
  filter(str_detect(form, "^[A-ZÄÖÜ][[:alpha:]]+$")) |>
  select(lemma, form) |>
  bind_rows(nouns |> transmute(lemma, form = lemma)) |>
  distinct(lemma, form) |>
  inner_join(nouns, by = "lemma") |>
  group_by(form) |>
  mutate(
    is_headword = form == lemma,
    has_headword = any(is_headword),
    number_of_lemmas = n_distinct(lemma)
  ) |>
  filter(
    (has_headword & is_headword) |
      (!has_headword & number_of_lemmas == 1)
  ) |>
  select(lemma, form, gender) |>
  ungroup()
```

The dictionary also includes case forms in both singular and plural. I reshape
those columns into a surface-form-to-lemma map. Exact headwords take precedence;
other spellings are discarded when they point to more than one lemma. Every
retained form is therefore assigned to one canonical dictionary lemma;
occurrences of *Häuser*, for example, contribute to *Haus*. Plural-only nouns
remain under the lemma supplied by the dictionary.

The subtitle list has no part-of-speech tags. Capitalization gets us most of the
way to identifying nouns, but a sentence-initial *Ich* is still capitalized. A
conservative safeguard is to keep a candidate only when its capitalized count is
at least five times its lowercase count (or the lowercase spelling is absent).
That retains *Frau* and *Mädchen*, whose lowercase occurrences are mostly noise,
while rejecting *Ich*, *Du*, and *Ist*. It also loses genuinely ambiguous nouns
such as *Leben*, but avoids assigning millions of pronoun and verb tokens to rare
metalinguistic nouns.

There is one residual problem that a context-free frequency list cannot solve:
some dictionary headwords are also overwhelmingly used as interjections or
function words in subtitles. I exclude a short, explicit list of those
homographs. This is intentionally auditable rather than a claim that *Oh* or
*Aber* can never be used nominally. A stronger follow-up would estimate noun use
from POS-tagged subtitle sentences instead of assigning every occurrence of a
spelling to its dictionary noun entry.

```{r}
#| label: clean-frequencies

read_frequency_page <- function(path) {
  lines <- read_lines(path, progress = FALSE)
  opening_brackets <- str_locate(lines, fixed("[["))[, 1]

  tibble(
    frequency = parse_number(lines),
    word = str_sub(lines, opening_brackets + 2L, -3L)
  ) |>
    filter(
      !is.na(frequency),
      str_detect(word, "^[A-Za-zÄÖÜäöüß]+$")
    )
}

subtitle_frequency <- map_dfr(subtitle_paths, read_frequency_page)

lowercase_frequency <- subtitle_frequency |>
  filter(word == str_to_lower(word, locale = "de")) |>
  transmute(
    lowercase_spelling = word,
    lowercase_frequency = frequency
  )

dominant_non_noun_uses <- c(
  "Ach", "Ah", "Aha", "Äh", "Ähm", "Au", "Bah", "Buh", "Danke", "Hallo",
  "He", "Hey", "Hi", "Hm", "Hmmm", "Ho", "Huch", "Hui", "Hurra", "Ja",
  "Nein", "Oh", "Oha", "Okay", "Psst", "Tja", "Ups", "Wow",
  "Aber", "Als", "Bitte", "Da", "Das", "Der", "Die", "Du", "Ein", "Er",
  "Es", "Ich", "Ihr", "Ist", "Los", "Man", "Mit", "Nach", "Sie", "So",
  "Und", "Vom", "Von", "Was", "Wer", "Wie", "Wir", "Zu"
)

capitalized_frequency <- subtitle_frequency |>
  filter(str_detect(word, "^[A-ZÄÖÜ]")) |>
  mutate(lowercase_spelling = str_to_lower(word, locale = "de")) |>
  left_join(lowercase_frequency, by = "lowercase_spelling") |>
  filter(
    is.na(lowercase_frequency) |
      frequency >= 5 * lowercase_frequency
  )

excluded_surface_frequency <- capitalized_frequency |>
  inner_join(noun_forms, by = c("word" = "form")) |>
  filter(word %in% dominant_non_noun_uses)

noun_surface_frequency <- capitalized_frequency |>
  filter(!word %in% dominant_non_noun_uses) |>
  inner_join(noun_forms, by = c("word" = "form")) |>
  transmute(
    lemma,
    gender,
    form = word,
    frequency
  )

noun_frequency <- noun_surface_frequency |>
  group_by(lemma, gender) |>
  summarise(
    frequency = sum(frequency),
    matched_forms = n_distinct(form),
    .groups = "drop"
  ) |>
  arrange(desc(frequency))

stopifnot(
  nrow(noun_forms) == n_distinct(noun_forms$form),
  nrow(noun_frequency) == n_distinct(noun_frequency$lemma)
)

noun_data <- noun_frequency

tibble(
  measure = c(
    "noun lemmas",
    "distinct inflected forms matched",
    "matched subtitle tokens",
    "explicitly excluded non-noun spellings"
  ),
  value = c(
    nrow(noun_data),
    n_distinct(noun_surface_frequency$form),
    sum(noun_data$frequency),
    n_distinct(excluded_surface_frequency$word)
  )
) |>
  kable(format.args = list(big.mark = ","))
```

The resulting `noun_frequency` table has one row per lemma. `matched_forms` makes
the singular/plural aggregation visible rather than hiding it inside the join.

```{r}
#| label: noun-frequency-preview

noun_frequency |>
  select(lemma, gender, matched_forms, frequency) |>
  slice_head(n = 12) |>
  kable(
    col.names = c("Lemma", "Gender", "Matched forms", "Frequency"),
    format.args = list(big.mark = ","),
    align = c("l", "l", "r", "r")
  )
```

There are two sensible scores. **Type accuracy** gives every dictionary entry one
vote. **Usage-weighted accuracy** gives a noun with frequency $f_i$ exactly
$f_i$ votes:

$$
  \operatorname{accuracy}_{\mathrm{usage}}
  = \frac{\sum_i f_i\,\mathbb{1}(\hat y_i = y_i)}{\sum_i f_i}.
$$

The second is the primary score because it better reflects how often a learner
would actually need the rule. I will still report both so that a handful of very
frequent words cannot silently define success.

## Compare the gender distributions

Before fitting anything, it is useful to ask what the class balance looks like.
Fedden et al. report two relevant reference distributions. Their cleaned database
of 30,576 noun types is 36% masculine, 45% feminine, and 19% neuter (their Table
1). They also reproduce CELEX token counts of 35%, 42%, and 23%, respectively
(their Table 3).

First, here is the comparison when every noun lemma gets one vote.

```{r}
#| label: type-distribution

paper_type_distribution <- tribble(
  ~gender, ~paper_noun_types, ~paper_type_share,
  "der", 10866, 0.36,
  "die", 13766, 0.45,
  "das",  5944, 0.19
) |>
  mutate(gender = factor(gender, levels = c("der", "die", "das")))

type_distribution <- noun_data |>
  count(gender, name = "our_noun_types") |>
  mutate(our_type_share = our_noun_types / sum(our_noun_types)) |>
  left_join(paper_type_distribution, by = "gender") |>
  mutate(
    gender = recode(
      as.character(gender),
      der = "Masculine (der)",
      die = "Feminine (die)",
      das = "Neuter (das)"
    ),
    across(ends_with("share"), function(x) percent(x, accuracy = 0.1))
  )

type_distribution |>
  kable(
    col.names = c(
      "Gender", "Our noun types", "Our share",
      "Paper noun types", "Paper share"
    ),
    format.args = list(big.mark = ",")
  )
```

Next, each noun is weighted by the number of matched subtitle occurrences. The
counts in the CELEX comparison are included for scale, although they come from a
different corpus.

```{r}
#| label: token-distribution

paper_token_distribution <- tribble(
  ~gender, ~celex_tokens, ~celex_token_share,
  "der", 380770, 0.35,
  "die", 447048, 0.42,
  "das", 248147, 0.23
) |>
  mutate(gender = factor(gender, levels = c("der", "die", "das")))

token_distribution <- noun_data |>
  count(gender, wt = frequency, name = "subtitle_occurrences") |>
  mutate(
    subtitle_weighted_share = subtitle_occurrences /
      sum(subtitle_occurrences)
  ) |>
  left_join(paper_token_distribution, by = "gender") |>
  mutate(
    gender = recode(
      as.character(gender),
      der = "Masculine (der)",
      die = "Feminine (die)",
      das = "Neuter (das)"
    ),
    across(ends_with("share"), function(x) percent(x, accuracy = 0.1))
  )

token_distribution |>
  kable(
    col.names = c(
      "Gender", "Matched subtitle occurrences", "Our weighted share",
      "CELEX tokens", "CELEX share"
    ),
    format.args = list(big.mark = ",")
  )
```

Our subtitle-filtered sample is more masculine and less feminine than the paper's
database, under both counting schemes. That is not necessarily a contradiction:
the authors' main point in this part of the paper is that gender proportions depend
on the corpus and the frequency range. Our counts cover dictionary forms that can
be assigned to a single lemma after the spelling filters, rather than POS-tagged
noun tokens, while the paper's CELEX token row is not subtitle-based. This
comparison is therefore a diagnostic, not a replication.

## Memorize the costly exceptions

The rule is meant to help with unfamiliar nouns, not replace vocabulary a learner
already knows. Otherwise a mistake on an extremely common word such as *Frau*,
*Mädchen*, or *Junge* dominates the usage-weighted loss even though a student is
unlikely to consult a gender rule for it.

But there is no need to memorize a frequent noun if the compact tree already gets
it right. In a preliminary pass, the tree correctly classified 51 of the original
100 most frequent nouns. I keep its 49 mistakes and add *Telefon*, the next most
frequent mistake, to obtain a round 50-word exception core. The list is frozen
before refitting the tree below.

This is now an exception list rather than a general beginner vocabulary. It is a
better match for the practical task: spend memory on the frequent words for which
the rule is unhelpful. Because the same dataset informed this design choice, the
combined score remains an exploratory estimate; a separate subtitle corpus would
be the right final test.

```{r}
#| label: core-vocabulary

core_lemmas <- c(
  "Angst", "Arbeit", "Art", "Auge", "Auto",
  "Baby", "Bett", "Bild", "Blut", "Buch",
  "Ding", "Ende", "Frau", "Freundin", "Gefühl",
  "Geld", "Gesicht", "Glück", "Hand", "Haus",
  "Herz", "Jahr", "Junge", "Kind", "Land",
  "Mama", "Miss", "Musik", "Mutter", "Mädchen",
  "Nachricht", "Nacht", "Name", "Nummer", "Polizei",
  "Problem", "Schiff", "Schuld", "Schwester", "Spiel",
  "Stadt", "Telefon", "Tochter", "Tür", "Uhr",
  "Wasser", "Welt", "Wort", "Zeit", "Zimmer"
)

core_vocabulary_size <- length(core_lemmas)

core_vocabulary <- noun_data |>
  filter(lemma %in% core_lemmas)

stopifnot(
  core_vocabulary_size == 50L,
  nrow(core_vocabulary) == core_vocabulary_size,
  setequal(core_vocabulary$lemma, core_lemmas)
)

rule_noun_data <- noun_data |>
  anti_join(core_vocabulary, by = "lemma")

tibble(
  vocabulary = c("Memorized exceptions", "Needs a rule"),
  noun_lemmas = c(nrow(core_vocabulary), nrow(rule_noun_data)),
  subtitle_tokens = c(
    sum(core_vocabulary$frequency),
    sum(rule_noun_data$frequency)
  )
) |>
  mutate(
    token_share = subtitle_tokens / sum(subtitle_tokens),
    token_share = percent(token_share, accuracy = 0.1)
  ) |>
  kable(
    col.names = c(
      "Vocabulary", "Noun lemmas", "Subtitle tokens", "Token share"
    ),
    format.args = list(big.mark = ",")
  )
```

All tree, baseline, and error scores below are conditional on a noun falling
outside this core. Later I also report a combined learner-system score that treats
the memorized exceptions as correct.

Here is the complete exception core, alphabetized by noun. The article is part of
each entry because that is what must be memorized.

```{r}
#| label: core-vocabulary-list

core_rows_per_column <- ceiling(nrow(core_vocabulary) / 4)

core_vocabulary |>
  arrange(str_to_lower(lemma, locale = "de")) |>
  transmute(
    entry = paste(gender, lemma),
    position = row_number(),
    row = (position - 1L) %% core_rows_per_column + 1L,
    column = paste0("column_", (position - 1L) %/% core_rows_per_column + 1L)
  ) |>
  select(row, column, entry) |>
  pivot_wider(names_from = column, values_from = entry) |>
  select(-row) |>
  kable(
    col.names = rep("Memorized exceptions", 4),
    align = rep("l", 4)
  )
```

[Download the printable exception core and decision tree (PDF)](/output/pdf/german-noun-gender-core-and-tree.pdf){.btn .btn-primary download="german-noun-gender-core-and-tree.pdf"}

## Turn spelling into learnable questions

The spelling features are not arbitrary character fragments. They are a short,
predeclared set of endings, beginnings, and word shapes that a learner could
plausibly recognize. This is where I add the proposed rules for *ge-*, *kn-*,
*dr-*, *tr-*, *-ft*, *-xt*, *-ur*, *-et*, and *-ett*. The tree is free to ignore
rules that do not improve held-out predictions.

```{r}
#| label: spelling-features

suffixes <- c(
  e = "e", chen = "chen", lein = "lein", ung = "ung",
  heit = "heit", keit = "keit", schaft = "schaft", ei = "ei",
  ie = "ie", ik = "ik", ion = "ion", taet = "tät",
  enz = "enz", anz = "anz", ur = "ur", `in` = "in", age = "age",
  er = "er", en = "en", el = "el", ich = "ich", ig = "ig",
  ling = "ling", ismus = "ismus", or = "or", us = "us",
  ment = "ment", um = "um", ma = "ma", o = "o"
)

count_written_syllables <- function(words) {
  words |>
    str_to_lower(locale = "de") |>
    str_replace_all("äu|au|ei|ai|eu|ie", "a_") |>
    str_count("[aeiouyäöü]+")
}

make_spelling_features <- function(words) {
  lowercase_words <- str_to_lower(words, locale = "de")

  suffixes |>
    map(~ str_ends(lowercase_words, fixed(.x))) |>
    set_names(paste0("ends_", names(suffixes))) |>
    as_tibble() |>
    mutate(
      starts_ge = str_starts(lowercase_words, fixed("ge")),
      starts_kn = str_starts(lowercase_words, fixed("kn")),
      starts_dr_or_tr = str_detect(lowercase_words, "^(dr|tr)"),
      one_written_syllable = count_written_syllables(words) == 1,
      ends_ft_or_xt = str_detect(lowercase_words, "(ft|xt)$"),
      ends_et_or_ett = str_detect(lowercase_words, "(et|ett)$")
    )
}
```

`one_written_syllable` is necessarily an orthographic approximation. It counts
vowel groups after collapsing common German diphthongs; it is useful for a toy
learner's rule, but it is not a phonetic syllabifier.

The remaining two rules need a verb lexicon. An exact infinitive spelling is a
proxy for a nominalized infinitive, while an exact match to an infinitive with
its final *-en* or *-n* removed is a proxy for a zero-derived verbal noun. These
tests catch examples such as *Essen* and *Kauf*, but they also flag homographs
such as *Rasen*. The classifier must decide whether the tendencies survive those
exceptions.

```{r}
#| label: verbal-features

verbs <- read_csv(
  verb_path,
  col_types = cols(.default = col_character()),
  show_col_types = FALSE
) |>
  transmute(
    infinitive = str_to_lower(Infinitive, locale = "de")
  ) |>
  filter(str_detect(infinitive, "^[a-zäöüß]+$")) |>
  distinct()

verb_infinitives <- verbs$infinitive
verb_stems <- verbs |>
  transmute(stem = str_remove(infinitive, "(?:en|n)$")) |>
  filter(str_length(stem) >= 3) |>
  distinct() |>
  pull(stem)

make_verbal_features <- function(words) {
  lowercase_words <- str_to_lower(words, locale = "de")

  tibble(
    is_infinitive = lowercase_words %in% verb_infinitives,
    has_verbal_base = lowercase_words %in% verb_stems
  )
}
```

## Add a small semantic layer

German also has semantic regularities. Weekdays, months, and the canonical seasons
are masculine, for example. The
[IDS grammar](https://grammis.ids-mannheim.de/systematische-grammatik/2275)
also discusses regularities for people, directions, colours, and other classes,
while stressing their exceptions and limited coverage.

The dictionary has no ontology, so I annotate a few small, auditable sets by hand.
These flags describe meaning, not the answer. That distinction matters for words
such as *Mädchen*: it denotes a female person, yet the diminutive ending makes the
noun neuter.

```{r}
#| label: semantic-features

weekdays <- c(
  "Montag", "Dienstag", "Mittwoch", "Donnerstag", "Freitag",
  "Samstag", "Sonnabend", "Sonntag"
)
months <- c(
  "Januar", "Februar", "März", "April", "Mai", "Juni", "Juli",
  "August", "September", "Oktober", "November", "Dezember"
)
seasons <- c("Frühling", "Sommer", "Herbst", "Winter")
directions <- c("Norden", "Süden", "Osten", "Westen")
colours <- c(
  "Rot", "Blau", "Grün", "Gelb", "Schwarz", "Weiß", "Braun",
  "Grau", "Rosa", "Violett"
)
female_people <- c(
  "Frau", "Mutter", "Tochter", "Schwester", "Tante", "Oma",
  "Großmutter", "Ehefrau", "Freundin", "Königin", "Prinzessin",
  "Dame", "Braut", "Witwe", "Nonne", "Hebamme", "Mädchen",
  "Fräulein"
)
male_people <- c(
  "Mann", "Herr", "Vater", "Sohn", "Bruder", "Onkel", "Opa",
  "Großvater", "Ehemann", "Freund", "König", "Prinz", "Kaiser",
  "Bräutigam", "Witwer", "Mönch", "Junge", "Männchen"
)

semantic_sets <- list(
  calendar = c(weekdays, months, seasons),
  direction = directions,
  colour = colours,
  female = female_people,
  male = male_people
)

model_data <- rule_noun_data |>
  bind_cols(
    make_spelling_features(rule_noun_data$lemma),
    make_verbal_features(rule_noun_data$lemma)
  ) |>
  mutate(
    meaning_calendar = lemma %in% semantic_sets$calendar,
    meaning_direction = lemma %in% semantic_sets$direction,
    meaning_colour = lemma %in% semantic_sets$colour,
    meaning_female = lemma %in% semantic_sets$female,
    meaning_male = lemma %in% semantic_sets$male
  )

feature_names <- c(
  paste0("ends_", names(suffixes)),
  "starts_ge", "starts_kn", "starts_dr_or_tr",
  "one_written_syllable", "ends_ft_or_xt", "ends_et_or_ett",
  "is_infinitive", "has_verbal_base",
  "meaning_calendar", "meaning_direction", "meaning_colour",
  "meaning_female", "meaning_male"
)

semantic_summary <- imap_dfr(semantic_sets, function(words, semantic_class) {
  rows <- filter(model_data, lemma %in% words)

  if (nrow(rows) == 0) {
    return(tibble(
      semantic_class,
      nouns = 0L,
      subtitle_tokens = 0,
      majority_article = NA_character_,
      weighted_purity = NA_real_
    ))
  }

  article_counts <- count(rows, gender, wt = frequency, name = "weight")

  tibble(
    semantic_class,
    nouns = nrow(rows),
    subtitle_tokens = sum(rows$frequency),
    majority_article = as.character(
      article_counts$gender[which.max(article_counts$weight)]
    ),
    weighted_purity = max(article_counts$weight) / sum(article_counts$weight)
  )
}) |>
  filter(nouns > 0)

semantic_summary |>
  mutate(weighted_purity = percent(weighted_purity, accuracy = 0.1)) |>
  kable(
    col.names = c(
      "Semantic class", "Nouns", "Subtitle tokens", "Majority article",
      "Weighted purity"
    ),
    format.args = list(big.mark = ",")
  )
```

This semantic annotation is intentionally small. An obvious extension would be to
replace the hand-built lists with a reproducible ontology, while keeping the
learner-facing feature names at this level of abstraction.

## Check the added rules

Before allowing the tree to combine the new features, I can test each tendency
on its own. “Usage-weighted hit rate” answers: among subtitle occurrences of
nouns matching this rule, what proportion has the proposed article? It is a
measure of a one-sided rule's reliability, not the accuracy of a complete
three-way classifier.

```{r}
#| label: added-rule-summary

weighted_accuracy <- function(truth, estimate, weight) {
  sum(weight * (truth == estimate)) / sum(weight)
}

added_rules <- tribble(
  ~feature, ~rule, ~predicted_article,
  "starts_ge", "starts with ge-", "das",
  "is_infinitive", "spelling is a verb infinitive", "das",
  "has_verbal_base", "spelling matches a verb stem", "der",
  "starts_kn", "starts with kn-", "der",
  "one_written_syllable", "one written syllable", "der",
  "starts_dr_or_tr", "starts with dr- or tr-", "der",
  "ends_ft_or_xt", "ends with -ft or -xt", "die",
  "ends_ur", "ends with -ur", "die",
  "ends_et_or_ett", "ends with -et or -ett", "das"
)

rule_summary <- pmap_dfr(
  added_rules,
  function(feature, rule, predicted_article) {
    matches <- model_data |>
      filter(.data[[feature]])

    tibble(
      rule,
      predicted_article,
      noun_lemmas = nrow(matches),
      subtitle_tokens = sum(matches$frequency),
      type_hit_rate = mean(matches$gender == predicted_article),
      usage_weighted_hit_rate = weighted_accuracy(
        matches$gender,
        factor(predicted_article, levels = levels(matches$gender)),
        matches$frequency
      )
    )
  }
)

rule_summary |>
  mutate(
    across(ends_with("hit_rate"), function(x) percent(x, accuracy = 0.1))
  ) |>
  kable(
    col.names = c(
      "Rule", "Predicted article", "Noun lemmas", "Subtitle tokens",
      "Type hit rate", "Usage-weighted hit rate"
    ),
    format.args = list(big.mark = ",")
  )
```

These rows also make weak or corpus-sensitive rules visible. Nothing is forced
into the final tree merely because it appeared on the candidate list.

## Start with one rule

The opening classifier is the folk rule proposed at the start: if a noun ends in
*-e*, predict *die*; otherwise predict *der*.

```{r}
#| label: baseline

baseline_prediction <- factor(
  if_else(model_data$ends_e, "die", "der"),
  levels = levels(model_data$gender)
)

baseline_scores <- tibble(
  model = "ends in -e → die; otherwise → der",
  type_accuracy = mean(baseline_prediction == model_data$gender),
  usage_weighted_accuracy = weighted_accuracy(
    model_data$gender,
    baseline_prediction,
    model_data$frequency
  )
)

baseline_scores |>
  mutate(across(ends_with("accuracy"), function(x) percent(x, accuracy = 0.1))) |>
  kable()
```

## Cross-validate trees of increasing size

To prevent the displayed result from being an in-sample score, I assign each
non-core lemma to one of ten folds. Every candidate tree is trained on nine folds
and predicts the tenth. The raw subtitle count is used as its case weight during
training as well as in the primary evaluation metric.

Tree size is measured in terminal leaves. I allow at most nine leaves for the
eventual learner's tree, but fit budgets up to sixteen to show what extra complexity
would buy.

```{r}
#| label: cross-validation

set.seed(20260904)
model_data <- model_data |>
  group_by(gender) |>
  mutate(fold = sample(rep(1:10, length.out = n()))) |>
  ungroup()

tree_formula <- reformulate(feature_names, response = "gender")

grow_tree <- function(data) {
  rpart(
    tree_formula,
    data = data,
    weights = frequency,
    method = "class",
    parms = list(split = "information"),
    control = rpart.control(
      cp = 0,
      minsplit = 30,
      minbucket = 10,
      maxdepth = 8,
      maxcompete = 0,
      maxsurrogate = 0,
      usesurrogate = 0,
      xval = 0
    )
  )
}

prune_to_budget <- function(tree, maximum_leaves) {
  path <- as_tibble(tree$cptable)
  selected <- path |>
    filter(nsplit <= maximum_leaves - 1L) |>
    slice_max(nsplit, n = 1, with_ties = FALSE)

  prune(tree, cp = selected$CP)
}

leaf_budgets <- 2:16

cv_predictions <- map_dfr(1:10, function(held_out_fold) {
  training <- filter(model_data, fold != held_out_fold)
  testing <- filter(model_data, fold == held_out_fold)
  full_tree <- grow_tree(training)

  map_dfr(leaf_budgets, function(budget) {
    candidate <- prune_to_budget(full_tree, budget)
    prediction <- predict(candidate, testing, type = "class")

    transmute(
      testing,
      fold = held_out_fold,
      leaf_budget = budget,
      actual_leaves = sum(candidate$frame$var == "<leaf>"),
      lemma,
      frequency,
      truth = gender,
      prediction
    )
  })
})

cv_summary <- cv_predictions |>
  group_by(leaf_budget) |>
  summarise(
    actual_leaves = median(actual_leaves),
    type_accuracy = mean(truth == prediction),
    usage_weighted_accuracy = weighted_accuracy(
      truth,
      prediction,
      frequency
    ),
    .groups = "drop"
  ) |>
  arrange(leaf_budget) |>
  mutate(
    previous_best = lag(cummax(usage_weighted_accuracy), default = -Inf),
    on_frontier = usage_weighted_accuracy > previous_best
  )
```

## Plot the accuracy--complexity frontier

A point is on the frontier when no smaller tree budget achieved equal or better
usage-weighted accuracy. The horizontal 96% line is deliberately aspirational: it
comes from a different study, feature set, and evaluation protocol.

```{r}
#| label: pareto-frontier
#| fig-cap: "Out-of-fold accuracy versus the maximum number of terminal leaves. Coloured points form the empirical Pareto frontier."
#| fig-width: 8
#| fig-height: 5

ggplot(cv_summary, aes(leaf_budget, usage_weighted_accuracy)) +
  geom_hline(
    yintercept = baseline_scores$usage_weighted_accuracy,
    colour = "grey55",
    linetype = "dotted"
  ) +
  geom_hline(yintercept = 0.96, colour = "grey45", linetype = "dashed") +
  geom_line(colour = "grey75") +
  geom_point(colour = "grey65", size = 2.5) +
  geom_line(
    data = filter(cv_summary, on_frontier),
    colour = "#0072B2",
    linewidth = 1
  ) +
  geom_point(
    data = filter(cv_summary, on_frontier),
    colour = "#0072B2",
    size = 3
  ) +
  scale_x_continuous(breaks = leaf_budgets) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  labs(
    x = "Maximum number of leaves",
    y = "Usage-weighted accuracy outside exception core",
    caption = "Dotted: one-rule baseline. Dashed: 96% rich-model reference."
  )
```

The two evaluation measures need not prefer the same tree. This table keeps that
trade-off visible.

```{r}
#| label: accuracy-table

cv_summary |>
  select(leaf_budget, actual_leaves, type_accuracy, usage_weighted_accuracy) |>
  mutate(across(ends_with("accuracy"), function(x) percent(x, accuracy = 0.1))) |>
  kable(
    col.names = c(
      "Leaf budget", "Median leaves", "Type accuracy",
      "Usage-weighted accuracy"
    )
  )
```

## Choose and draw one learnable tree

I define “learnable” before looking at the answer: no more than nine leaves and
no exact-word questions. Among those candidates I choose the smallest tree within
0.25 percentage points of the best usage-weighted score. The tolerance prevents
a barely measurable gain from adding several more rules.

```{r}
#| label: choose-tree

learnable_candidates <- cv_summary |>
  filter(leaf_budget <= 9)

best_learnable_accuracy <- max(
  learnable_candidates$usage_weighted_accuracy
)

chosen_budget <- learnable_candidates |>
  filter(
    usage_weighted_accuracy >= best_learnable_accuracy - 0.0025,
    on_frontier
  ) |>
  slice_min(leaf_budget, n = 1, with_ties = FALSE) |>
  pull(leaf_budget)

full_tree <- grow_tree(model_data)
final_tree <- full_tree |>
  prune_to_budget(chosen_budget)

chosen_tree_size <- cv_summary |>
  filter(leaf_budget == chosen_budget) |>
  transmute(
    leaf_budget,
    median_cv_leaves = actual_leaves,
    final_tree_leaves = sum(final_tree$frame$var == "<leaf>")
  )

chosen_tree_size |>
  kable(
    col.names = c(
      "Leaf budget", "Median CV leaves", "Final tree leaves"
    )
  )

chosen_predictions <- cv_predictions |>
  filter(leaf_budget == chosen_budget)

chosen_scores <- tibble(
  evaluation = c(
    "Tree on nouns outside the exception core",
    "Exception core memorized plus tree"
  ),
  type_accuracy = c(
    mean(chosen_predictions$truth == chosen_predictions$prediction),
    (
      nrow(core_vocabulary) +
        sum(chosen_predictions$truth == chosen_predictions$prediction)
    ) / nrow(noun_data)
  ),
  usage_weighted_accuracy = c(
    weighted_accuracy(
      chosen_predictions$truth,
      chosen_predictions$prediction,
      chosen_predictions$frequency
    ),
    (
      sum(core_vocabulary$frequency) +
        sum(
          chosen_predictions$frequency *
            (chosen_predictions$truth == chosen_predictions$prediction)
        )
    ) / sum(noun_data$frequency)
  )
)

chosen_scores |>
  mutate(across(ends_with("accuracy"), function(x) percent(x, accuracy = 0.1))) |>
  kable(
    col.names = c(
      "Evaluation", "Type accuracy", "Usage-weighted accuracy"
    )
  )
```

The leaf budget is a ceiling, not a promise. Cost-complexity pruning offers only
the subtree sizes on its pruning path, so the tree fitted to all the data can
have fewer leaves than the selected maximum. The table distinguishes its actual
size from the median size seen in cross-validation.

```{r}
#| label: final-tree
#| fig-cap: "The final frequency-weighted decision tree for nouns outside the exception core. Consecutive questions with the same outcome are combined into one OR rule."
#| fig-width: 9
#| fig-height: 5.5

question_by_feature <- c(
  setNames(
    sprintf("Ends in -%s?", unname(suffixes)),
    paste0("ends_", names(suffixes))
  ),
  starts_ge = "Starts with ge-?",
  starts_kn = "Starts with kn-?",
  starts_dr_or_tr = "Starts with dr- or tr-?",
  one_written_syllable = "One written syllable?",
  ends_ft_or_xt = "Ends in -ft or -xt?",
  ends_et_or_ett = "Ends in -et or -ett?",
  is_infinitive = "Also a verb infinitive?",
  has_verbal_base = "Has a verbal base?",
  meaning_calendar = "A day, month, or season?",
  meaning_direction = "A compass direction?",
  meaning_colour = "A colour?",
  meaning_female = "A female person?",
  meaning_male = "A male person?"
)

tree_frame <- final_tree$frame
article_levels <- levels(model_data$gender)

is_leaf <- function(node) {
  tree_frame[as.character(node), "var"] == "<leaf>"
}

node_article <- function(node) {
  article_levels[tree_frame[as.character(node), "yval"]]
}

# Logical predictors are stored by rpart as 0/1 numeric splits. The sign of
# ncat records whether TRUE goes left or right in the fitted tree.
internal_nodes <- row.names(tree_frame)[tree_frame$var != "<leaf>"]
split_direction <- setNames(
  sign(final_tree$splits[, "ncat"]),
  internal_nodes
)

stopifnot(all(final_tree$splits[, "index"] == 0.5))

decision_rows <- list()
current_node <- 1L
step <- 1L

repeat {
  feature <- tree_frame[as.character(current_node), "var"]
  left_node <- current_node * 2L
  right_node <- left_node + 1L
  true_goes_left <- split_direction[as.character(current_node)] > 0
  true_node <- if (true_goes_left) left_node else right_node
  false_node <- if (true_goes_left) right_node else left_node

  # The selected pruned tree is a decision list: TRUE predicts immediately,
  # while FALSE either asks the next question or reaches the default leaf.
  stopifnot(!is_leaf(current_node), is_leaf(true_node))

  decision_rows[[step]] <- tibble(
    step,
    feature,
    question = unname(question_by_feature[feature]),
    yes_article = node_article(true_node),
    no_article = if (is_leaf(false_node)) node_article(false_node) else NA_character_
  )

  if (is_leaf(false_node)) {
    break
  }

  current_node <- false_node
  step <- step + 1L
}

raw_decision_list <- bind_rows(decision_rows)

collapse_with_or <- function(values) {
  if (length(values) == 1L) {
    return(values)
  }

  if (length(values) == 2L) {
    return(paste(values, collapse = " or "))
  }

  paste0(paste(head(values, -1L), collapse = ", "), ", or ", tail(values, 1L))
}

ending_by_feature <- c(
  setNames(paste0("-", unname(suffixes)), paste0("ends_", names(suffixes))),
  ends_ft_or_xt = "-ft/-xt",
  ends_et_or_ett = "-et/-ett"
)

combine_questions <- function(features, questions) {
  if (all(features %in% names(ending_by_feature))) {
    endings <- unname(ending_by_feature[features])
    return(paste0("Ends in ", collapse_with_or(endings), "?"))
  }

  if (length(questions) == 1L) {
    return(questions)
  }

  collapse_with_or(questions)
}

# Consecutive questions with the same Yes outcome are logically one OR rule.
# Collapsing them makes the learner-facing diagram much less repetitive while
# preserving the fitted tree's predictions.
decision_list <- raw_decision_list |>
  mutate(
    display_group = cumsum(
      yes_article != lag(yes_article, default = first(yes_article))
    )
  ) |>
  group_by(display_group) |>
  summarise(
    step = min(step),
    question = combine_questions(feature, question),
    feature = paste(feature, collapse = "|"),
    yes_article = first(yes_article),
    no_article = if (any(!is.na(no_article))) {
      first(na.omit(no_article))
    } else {
      NA_character_
    },
    .groups = "drop"
  ) |>
  mutate(
    x = 0,
    y = rev(seq_len(n()))
  )

yes_leaves <- decision_list |>
  transmute(x = 2.35, y, article = yes_article)

default_leaf <- decision_list |>
  filter(!is.na(no_article)) |>
  transmute(x = 0, y = 0, article = no_article)

vertical_branches <- decision_list |>
  transmute(
    x = 0,
    xend = 0,
    y = y - 0.27,
    yend = y - 0.73,
    label_x = 0.18,
    label_y = y - 0.5
  )

article_colours <- c(
  der = "#2563EB",
  die = "#DB2777",
  das = "#D97706"
)

ggplot() +
  geom_segment(
    data = decision_list,
    aes(x = 0.78, xend = 1.88, y = y, yend = y),
    colour = "#64748B",
    linewidth = 0.7,
    arrow = grid::arrow(length = grid::unit(0.09, "inches"), type = "closed")
  ) +
  geom_segment(
    data = vertical_branches,
    aes(x = x, xend = xend, y = y, yend = yend),
    colour = "#64748B",
    linewidth = 0.7,
    arrow = grid::arrow(length = grid::unit(0.09, "inches"), type = "closed")
  ) +
  geom_text(
    data = decision_list,
    aes(x = 1.32, y = y + 0.16, label = "Yes"),
    colour = "#475569",
    size = 3.5
  ) +
  geom_text(
    data = vertical_branches,
    aes(x = label_x, y = label_y, label = "No"),
    hjust = 0,
    colour = "#475569",
    size = 3.5
  ) +
  geom_label(
    data = decision_list,
    aes(x = x, y = y, label = question),
    fill = "#F8F4EA",
    colour = "#243447",
    linewidth = 0.35,
    label.padding = grid::unit(0.28, "lines"),
    label.r = grid::unit(0.18, "lines"),
    fontface = "bold",
    size = 4.1
  ) +
  geom_label(
    data = bind_rows(yes_leaves, default_leaf),
    aes(x = x, y = y, label = article, fill = article),
    colour = "white",
    linewidth = 0,
    label.padding = grid::unit(0.38, "lines"),
    label.r = grid::unit(0.6, "lines"),
    fontface = "bold",
    size = 4.5
  ) +
  annotate(
    "text",
    x = -0.34,
    y = 0,
    label = "otherwise",
    hjust = 1,
    colour = "#64748B",
    size = 3.4
  ) +
  scale_fill_manual(values = article_colours, guide = "none") +
  coord_cartesian(
    xlim = c(-1.1, 2.9),
    ylim = c(-0.5, nrow(decision_list) + 0.55),
    clip = "off"
  ) +
  labs(
    title = "A compact German noun-gender decision list",
    subtitle = "For a noun outside the exception core, follow No downward"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", colour = "#172033", size = 17),
    plot.subtitle = element_text(colour = "#64748B", margin = margin(b = 12)),
    plot.margin = margin(12, 24, 12, 24)
  )
```

Feature names translate directly into questions: `ends_chen` means “does the word
end in *-chen*?”, and `meaning_calendar` means “is it a weekday, month, or
canonical season?” Likewise, `is_infinitive` asks whether the spelling occurs in
the verb lexicon, and `one_written_syllable` uses the deliberately simple
vowel-group approximation described above. The resulting tree is the object to
critique as a learner: are its rules memorable, and is the accuracy gain worth
each additional branch?

One semantic feature now survives pruning: asking whether the noun denotes a male
person. The other semantic sets do not buy as much predictive accuracy as the
suffixes. A true linguistic generalization need not earn a place in a severely
pruned predictive tree.

Of the nine newly added tendencies, the *-ft/-xt* rule does enter the final tree.
The others remain useful learner-facing evidence in the individual-rule table,
but do not clear the accuracy threshold under the nine-leaf ceiling.

The absence of a *das* leaf is not a plotting error. Under an objective that
maximizes overall usage-weighted accuracy, the limited leaves are more profitably
spent identifying feminine endings and using *der* as the fallback. That is an
important failure mode of aggregate accuracy: if minimum performance on every
gender matters, the fitting objective would need class weights or an explicit
per-gender recall constraint.

### Why does the tree have only one continuing branch?

This shape is usually called a **decision list**: both outcomes exist at every
split, but one outcome receives an article immediately and only the other needs
another question. The learner-facing diagram combines consecutive questions that
have the same outcome. Read from the root, it says: endings in *-e* or *-ung* give
*die*; otherwise a male person gives *der*; otherwise the endings *-ion*, *-heit*,
*-keit*, *-ft*, and *-xt* give *die*; and the fallback is *der*.

`rpart` does not try to make a visually balanced tree. At each node it greedily
chooses the split with the greatest information gain. Because most predictors
here are sparse yes/no rules, a useful split tends to peel off a small,
high-purity group while leaving a large mixed group for the next rule. The
frequency weights make that behaviour stronger when the peeled-off nouns are
common. Splitting the already pure side would buy little accuracy and can also
run into the `minsplit` and `minbucket` requirements.

The larger `full_tree` shows that this is not a plotting constraint: before
pruning it contains nodes whose two outcomes both lead to further questions.
Cross-validation does not find their improvement large enough for the learner's
complexity budget, so cost-complexity pruning collapses them into leaves and
leaves the displayed decision list.

## Inspect the costly mistakes

Accuracy hides which errors matter. The following table uses out-of-fold
predictions and sorts mistakes by subtitle frequency, so it is also a practical
study list for further exceptions. By construction, memorized words such
as *Frau*, *Mädchen*, and *Junge* do not appear here.

```{r}
#| label: errors

cv_predictions |>
  filter(leaf_budget == chosen_budget, truth != prediction) |>
  arrange(desc(frequency)) |>
  select(lemma, truth, prediction, frequency) |>
  slice_head(n = 20) |>
  kable(
    col.names = c("Noun", "Correct article", "Predicted", "Frequency"),
    format.args = list(big.mark = ",")
  )
```

## What this experiment does not establish

This is a model of dictionary lemmas, not noun tokens in context. The conservative
capitalization filter removes many ambiguous words. The frequency aggregation
does include every recorded singular and plural case form that maps unambiguously
to a lemma, but the source still cannot tell a noun use from a sentence-initial
homograph. Subtitle dialogue is closer to everyday speech than a dictionary's
uniform type distribution, but translated dialogue is not identical to
spontaneous German conversation.

The split is also random by lemma. A stricter follow-up would group morphologically
related nouns or shared compound heads in the same fold, preventing near-duplicate
formations from appearing in both training and test data. Other useful sensitivity
checks would replace raw frequency by `log1p(frequency)`, vary the 50-word
exception budget or replace it with a CEFR vocabulary list, and expand the
semantic annotations with a real ontology.

Most importantly, 96% is a benchmark for a much richer statistical model. The
point of this post is to measure how much accuracy can be retained after imposing
a severe human-memory budget.
