---
title: "How many rules does it take to guess a German noun's gender?"
date: 2026-09-04
categories:
  - r
  - machine learning
  - german
execute:
  warning: false
  message: false
---

German noun gender is a nice case where the most accurate model is not necessarily
the most useful one. The object I want at the end of this analysis is a single
decision tree small enough for a learner to remember.

That gives the exercise two objectives:

1. predict the right article for nouns one is likely to encounter in practice;
2. use as few intelligible rules as possible.

The output will therefore be a Pareto frontier of predictive accuracy against tree
size, followed by one deliberately small tree. A flexible model reported by
[Fedden et al.](https://openresearch.surrey.ac.uk/esploro/outputs/journalArticle/Typology-meets-statistical-modeling-The-German/991154296202346)
reached about 96% accuracy with much richer phonological, semantic, derivational,
and inflectional information. I will use that number as a research-grade reference,
not as a theoretical upper bound or a directly comparable score.

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(knitr)
library(purrr)
library(readr)
library(rpart)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

theme_set(theme_minimal(base_size = 13))
```

## Download the dictionary and subtitle frequencies

The labels come from the
[german-nouns](https://github.com/gambolputty/german-nouns) export of German
Wiktionary. Its CSV contains roughly 100,000 entries, including gender and
inflected forms, and is published under CC BY-SA 4.0.

The usage weights come from Matthias Buchmeier's
[German subtitle frequency list](https://en.wiktionary.org/wiki/User:Matthias_Buchmeier#German_frequency_list).
It was made from 25,399,099 words of movie and television subtitles and preserves
letter case. The list is split into pages of 5,000 entries; the first ten pages are
enough for this experiment.

The following chunk downloads about 21 MB in total and caches it in this post's
`data` directory. That directory is ignored by Git.

```{r}
#| label: download-data

data_dir <- "data"
dir.create(data_dir, showWarnings = FALSE)

noun_url <- paste0(
  "https://raw.githubusercontent.com/gambolputty/german-nouns/",
  "main/german_nouns/nouns.csv"
)
noun_path <- file.path(data_dir, "nouns.csv")

if (!file.exists(noun_path)) {
  download.file(noun_url, noun_path, mode = "wb", quiet = TRUE)
}

starts <- seq(1, 45001, by = 5000)
ends <- starts + 4999
page_names <- sprintf("German_frequency_list-%d-%d", starts, ends)
subtitle_urls <- sprintf(
  paste0(
    "https://en.wiktionary.org/w/index.php?title=",
    "User:Matthias_Buchmeier/%s&action=raw"
  ),
  page_names
)
subtitle_paths <- file.path(data_dir, paste0(page_names, ".txt"))

walk2(subtitle_urls, subtitle_paths, function(url, path) {
  if (!file.exists(path)) {
    download.file(URLencode(url), path, mode = "wb", quiet = TRUE)
  }
})
```

For a fresh R installation, these are the required packages:

```r
install.packages(c(
  "dplyr", "ggplot2", "knitr", "purrr", "readr", "rpart",
  "scales", "stringr", "tibble", "tidyr"
))
```

## Build a frequency-weighted noun dictionary

The dictionary includes affixes, proper names, phrases, and nouns with more than
one possible gender. I keep ordinary one-word nouns whose Wiktionary record has
exactly one gender.

```{r}
#| label: clean-dictionary

gender_columns <- c("genus", paste("genus", 1:4))

nouns <- read_csv(
  noun_path,
  col_types = cols(.default = col_character()),
  show_col_types = FALSE
) |>
  filter(
    pos == "Substantiv",
    str_detect(lemma, "^[A-ZÄÖÜ][[:alpha:]]+$")
  ) |>
  select(lemma, all_of(gender_columns)) |>
  pivot_longer(
    cols = all_of(gender_columns),
    names_to = NULL,
    values_to = "gender"
  ) |>
  filter(gender %in% c("m", "f", "n")) |>
  distinct(lemma, gender) |>
  add_count(lemma, name = "number_of_genders") |>
  filter(number_of_genders == 1) |>
  transmute(
    lemma,
    gender = factor(
      recode(gender, m = "der", f = "die", n = "das"),
      levels = c("der", "die", "das")
    )
  )
```

The subtitle list has no part-of-speech tags. Capitalization gets us most of the
way to identifying nouns, but a sentence-initial *Ich* is still capitalized. A
conservative safeguard is to keep a candidate only when its capitalized count is
at least five times its lowercase count (or the lowercase spelling is absent).
That retains *Frau* and *Mädchen*, whose lowercase occurrences are mostly noise,
while rejecting *Ich*, *Du*, and *Ist*. It also loses genuinely ambiguous nouns
such as *Leben*, but avoids assigning millions of pronoun and verb tokens to rare
metalinguistic nouns.

```{r}
#| label: clean-frequencies

read_frequency_page <- function(path) {
  lines <- read_lines(path, progress = FALSE)
  opening_brackets <- str_locate(lines, fixed("[["))[, 1]

  tibble(
    frequency = parse_number(lines),
    word = str_sub(lines, opening_brackets + 2L, -3L)
  ) |>
    filter(
      !is.na(frequency),
      str_detect(word, "^[A-Za-zÄÖÜäöüß]+$")
    )
}

subtitle_frequency <- map_dfr(subtitle_paths, read_frequency_page)

lowercase_frequency <- subtitle_frequency |>
  filter(word == str_to_lower(word, locale = "de")) |>
  transmute(
    lowercase_spelling = word,
    lowercase_frequency = frequency
  )

noun_frequency <- subtitle_frequency |>
  filter(str_detect(word, "^[A-ZÄÖÜ]")) |>
  mutate(lowercase_spelling = str_to_lower(word, locale = "de")) |>
  left_join(lowercase_frequency, by = "lowercase_spelling") |>
  filter(
    is.na(lowercase_frequency) |
      frequency >= 5 * lowercase_frequency
  ) |>
  transmute(lemma = word, frequency)

noun_data <- nouns |>
  inner_join(noun_frequency, by = "lemma") |>
  arrange(desc(frequency))

tibble(
  measure = c("noun lemmas", "matched subtitle tokens"),
  value = c(nrow(noun_data), sum(noun_data$frequency))
) |>
  kable(format.args = list(big.mark = ","))
```

There are two sensible scores. **Type accuracy** gives every dictionary entry one
vote. **Usage-weighted accuracy** gives a noun with frequency (f_i) exactly
(f_i) votes:

\[
  \operatorname{accuracy}_{\mathrm{usage}}
  = \frac{\sum_i f_i\,\mathbb{1}(\hat y_i = y_i)}{\sum_i f_i}.
\]

The second is the primary score because it better reflects how often a learner
would actually need the rule. I will still report both so that a handful of very
frequent words cannot silently define success.

## Turn spelling into learnable questions

The spelling features are not arbitrary character fragments. They are a short,
predeclared set of endings that a learner could plausibly recognize. The tree is
free to ignore most of them.

```{r}
#| label: spelling-features

suffixes <- c(
  e = "e", chen = "chen", lein = "lein", ung = "ung",
  heit = "heit", keit = "keit", schaft = "schaft", ei = "ei",
  ie = "ie", ik = "ik", ion = "ion", taet = "tät",
  enz = "enz", anz = "anz", ur = "ur", `in` = "in", age = "age",
  er = "er", en = "en", el = "el", ich = "ich", ig = "ig",
  ling = "ling", ismus = "ismus", or = "or", us = "us",
  ment = "ment", um = "um", ma = "ma", o = "o"
)

make_suffix_features <- function(words) {
  features <- lapply(suffixes, function(suffix) {
    str_ends(str_to_lower(words, locale = "de"), fixed(suffix))
  })

  as_tibble(setNames(features, paste0("ends_", names(suffixes))))
}
```

## Add a small semantic layer

German also has semantic regularities. Weekdays, months, and the canonical seasons
are masculine, for example. The
[IDS grammar](https://grammis.ids-mannheim.de/systematische-grammatik/2275)
also discusses regularities for people, directions, colours, and other classes,
while stressing their exceptions and limited coverage.

The dictionary has no ontology, so I annotate a few small, auditable sets by hand.
These flags describe meaning, not the answer. That distinction matters for words
such as *Mädchen*: it denotes a female person, yet the diminutive ending makes the
noun neuter.

```{r}
#| label: semantic-features

weekdays <- c(
  "Montag", "Dienstag", "Mittwoch", "Donnerstag", "Freitag",
  "Samstag", "Sonnabend", "Sonntag"
)
months <- c(
  "Januar", "Februar", "März", "April", "Mai", "Juni", "Juli",
  "August", "September", "Oktober", "November", "Dezember"
)
seasons <- c("Frühling", "Sommer", "Herbst", "Winter")
directions <- c("Norden", "Süden", "Osten", "Westen")
colours <- c(
  "Rot", "Blau", "Grün", "Gelb", "Schwarz", "Weiß", "Braun",
  "Grau", "Rosa", "Violett"
)
female_people <- c(
  "Frau", "Mutter", "Tochter", "Schwester", "Tante", "Oma",
  "Großmutter", "Ehefrau", "Freundin", "Königin", "Prinzessin",
  "Dame", "Braut", "Witwe", "Nonne", "Hebamme", "Mädchen",
  "Fräulein"
)
male_people <- c(
  "Mann", "Herr", "Vater", "Sohn", "Bruder", "Onkel", "Opa",
  "Großvater", "Ehemann", "Freund", "König", "Prinz", "Kaiser",
  "Bräutigam", "Witwer", "Mönch", "Junge", "Männchen"
)

semantic_sets <- list(
  calendar = c(weekdays, months, seasons),
  direction = directions,
  colour = colours,
  female = female_people,
  male = male_people
)

model_data <- noun_data |>
  bind_cols(make_suffix_features(noun_data$lemma)) |>
  mutate(
    meaning_calendar = lemma %in% semantic_sets$calendar,
    meaning_direction = lemma %in% semantic_sets$direction,
    meaning_colour = lemma %in% semantic_sets$colour,
    meaning_female = lemma %in% semantic_sets$female,
    meaning_male = lemma %in% semantic_sets$male
  )

feature_names <- c(
  paste0("ends_", names(suffixes)),
  "meaning_calendar", "meaning_direction", "meaning_colour",
  "meaning_female", "meaning_male"
)

semantic_summary <- imap_dfr(semantic_sets, function(words, semantic_class) {
  rows <- filter(model_data, lemma %in% words)
  article_counts <- count(rows, gender, wt = frequency, name = "weight")

  tibble(
    semantic_class,
    nouns = nrow(rows),
    subtitle_tokens = sum(rows$frequency),
    majority_article = as.character(
      article_counts$gender[which.max(article_counts$weight)]
    ),
    weighted_purity = max(article_counts$weight) / sum(article_counts$weight)
  )
})

semantic_summary |>
  mutate(weighted_purity = percent(weighted_purity, accuracy = 0.1)) |>
  kable(
    col.names = c(
      "Semantic class", "Nouns", "Subtitle tokens", "Majority article",
      "Weighted purity"
    ),
    format.args = list(big.mark = ",")
  )
```

This semantic annotation is intentionally small. An obvious extension would be to
replace the hand-built lists with a reproducible ontology, while keeping the
learner-facing feature names at this level of abstraction.

## Start with one rule

The opening classifier is the folk rule proposed at the start: if a noun ends in
*-e*, predict *die*; otherwise predict *der*.

```{r}
#| label: baseline

weighted_accuracy <- function(truth, estimate, weight) {
  sum(weight * (truth == estimate)) / sum(weight)
}

baseline_prediction <- factor(
  if_else(model_data$ends_e, "die", "der"),
  levels = levels(model_data$gender)
)

baseline_scores <- tibble(
  model = "ends in -e → die; otherwise → der",
  type_accuracy = mean(baseline_prediction == model_data$gender),
  usage_weighted_accuracy = weighted_accuracy(
    model_data$gender,
    baseline_prediction,
    model_data$frequency
  )
)

baseline_scores |>
  mutate(across(ends_with("accuracy"), function(x) percent(x, accuracy = 0.1))) |>
  kable()
```

## Cross-validate trees of increasing size

To prevent the displayed result from being an in-sample score, I assign each lemma
to one of ten folds. Every candidate tree is trained on nine folds and predicts the
tenth. The raw subtitle count is used as its case weight during training as well as
in the primary evaluation metric.

Tree size is measured in terminal leaves. I allow at most eight leaves for the
eventual learner's tree, but fit budgets up to sixteen to show what extra complexity
would buy.

```{r}
#| label: cross-validation

set.seed(20260904)
model_data <- model_data |>
  group_by(gender) |>
  mutate(fold = sample(rep(1:10, length.out = n()))) |>
  ungroup()

tree_formula <- reformulate(feature_names, response = "gender")

grow_tree <- function(data) {
  rpart(
    tree_formula,
    data = data,
    weights = frequency,
    method = "class",
    parms = list(split = "information"),
    control = rpart.control(
      cp = 0,
      minsplit = 30,
      minbucket = 10,
      maxdepth = 8,
      maxcompete = 0,
      maxsurrogate = 0,
      usesurrogate = 0,
      xval = 0
    )
  )
}

prune_to_budget <- function(tree, maximum_leaves) {
  path <- as_tibble(tree$cptable)
  selected <- path |>
    filter(nsplit <= maximum_leaves - 1L) |>
    slice_max(nsplit, n = 1, with_ties = FALSE)

  prune(tree, cp = selected$CP)
}

leaf_budgets <- 2:16

cv_predictions <- map_dfr(1:10, function(held_out_fold) {
  training <- filter(model_data, fold != held_out_fold)
  testing <- filter(model_data, fold == held_out_fold)
  full_tree <- grow_tree(training)

  map_dfr(leaf_budgets, function(budget) {
    candidate <- prune_to_budget(full_tree, budget)
    prediction <- predict(candidate, testing, type = "class")

    transmute(
      testing,
      fold = held_out_fold,
      leaf_budget = budget,
      actual_leaves = sum(candidate$frame$var == "<leaf>"),
      lemma,
      frequency,
      truth = gender,
      prediction
    )
  })
})

cv_summary <- cv_predictions |>
  group_by(leaf_budget) |>
  summarise(
    actual_leaves = median(actual_leaves),
    type_accuracy = mean(truth == prediction),
    usage_weighted_accuracy = weighted_accuracy(
      truth,
      prediction,
      frequency
    ),
    .groups = "drop"
  ) |>
  arrange(leaf_budget) |>
  mutate(
    previous_best = lag(cummax(usage_weighted_accuracy), default = -Inf),
    on_frontier = usage_weighted_accuracy > previous_best
  )
```

## Plot the accuracy--complexity frontier

A point is on the frontier when no smaller tree budget achieved equal or better
usage-weighted accuracy. The horizontal 96% line is deliberately aspirational: it
comes from a different study, feature set, and evaluation protocol.

```{r}
#| label: pareto-frontier
#| fig-cap: "Out-of-fold accuracy versus the maximum number of terminal leaves. Coloured points form the empirical Pareto frontier."
#| fig-width: 8
#| fig-height: 5

ggplot(cv_summary, aes(leaf_budget, usage_weighted_accuracy)) +
  geom_hline(
    yintercept = baseline_scores$usage_weighted_accuracy,
    colour = "grey55",
    linetype = "dotted"
  ) +
  geom_hline(yintercept = 0.96, colour = "grey45", linetype = "dashed") +
  geom_line(colour = "grey75") +
  geom_point(colour = "grey65", size = 2.5) +
  geom_line(
    data = filter(cv_summary, on_frontier),
    colour = "#0072B2",
    linewidth = 1
  ) +
  geom_point(
    data = filter(cv_summary, on_frontier),
    colour = "#0072B2",
    size = 3
  ) +
  scale_x_continuous(breaks = leaf_budgets) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  labs(
    x = "Maximum number of leaves",
    y = "Usage-weighted accuracy",
    caption = "Dotted: one-rule baseline. Dashed: 96% rich-model reference."
  )
```

The two evaluation measures need not prefer the same tree. This table keeps that
trade-off visible.

```{r}
#| label: accuracy-table

cv_summary |>
  select(leaf_budget, actual_leaves, type_accuracy, usage_weighted_accuracy) |>
  mutate(across(ends_with("accuracy"), function(x) percent(x, accuracy = 0.1))) |>
  kable(
    col.names = c(
      "Leaf budget", "Median leaves", "Type accuracy",
      "Usage-weighted accuracy"
    )
  )
```

## Choose and draw one learnable tree

I define “learnable” before looking at the answer: no more than eight leaves and
no exact-word questions. Among
those candidates I choose the smallest tree within 0.25 percentage points of the
best usage-weighted score. The tolerance prevents a barely measurable gain from
doubling the number of rules.

```{r}
#| label: choose-tree

learnable_candidates <- cv_summary |>
  filter(leaf_budget <= 8)

best_learnable_accuracy <- max(
  learnable_candidates$usage_weighted_accuracy
)

chosen_budget <- learnable_candidates |>
  filter(
    usage_weighted_accuracy >= best_learnable_accuracy - 0.0025,
    on_frontier
  ) |>
  slice_min(leaf_budget, n = 1, with_ties = FALSE) |>
  pull(leaf_budget)

final_tree <- grow_tree(model_data) |>
  prune_to_budget(chosen_budget)

chosen_scores <- cv_summary |>
  filter(leaf_budget == chosen_budget) |>
  select(leaf_budget, actual_leaves, type_accuracy, usage_weighted_accuracy)

chosen_scores |>
  mutate(across(ends_with("accuracy"), function(x) percent(x, accuracy = 0.1))) |>
  kable()
```

```{r}
#| label: final-tree
#| fig-cap: "The final frequency-weighted decision tree. Each internal label is a yes/no question and each leaf is an article prediction."
#| fig-width: 10
#| fig-height: 6

plot(
  final_tree,
  uniform = TRUE,
  branch = 0.5,
  compress = FALSE,
  margin = 0.08
)
text(final_tree, use.n = FALSE, all = FALSE, pretty = 0, cex = 0.9)
```

Feature names translate directly into questions: `ends_chen` means “does the word
end in *-chen*?”, and `meaning_calendar` means “is it a weekday, month, or
canonical season?” The resulting tree is the object to critique as a learner: are
its rules memorable, and is the accuracy gain worth each additional branch?

The calendar feature is present in the dataset but does not appear in this small
tree. That is informative rather than a bug: all of the annotated calendar names
are masculine, and the default prediction on their spelling branch is already
*der*. A true linguistic generalization need not buy predictive accuracy. The
male- and female-person features do enter the fitted tree because they protect or
correct exceptions that suffixes alone would mishandle.

## Inspect the costly mistakes

Accuracy hides which errors matter. The following table uses out-of-fold
predictions and sorts mistakes by subtitle frequency, so it is also a practical
study list for exceptions to memorize.

```{r}
#| label: errors

cv_predictions |>
  filter(leaf_budget == chosen_budget, truth != prediction) |>
  arrange(desc(frequency)) |>
  select(lemma, truth, prediction, frequency) |>
  slice_head(n = 20) |>
  kable(
    col.names = c("Noun", "Correct article", "Predicted", "Frequency"),
    format.args = list(big.mark = ",")
  )
```

## What this experiment does not establish

This is a model of dictionary lemmas, not noun tokens in context. The conservative
capitalization filter removes many ambiguous words, and exact lemma counts miss
plural and declined forms. Subtitle dialogue is closer to everyday speech than a
dictionary's uniform type distribution, but translated dialogue is not identical
to spontaneous German conversation.

The split is also random by lemma. A stricter follow-up would group morphologically
related nouns or shared compound heads in the same fold, preventing near-duplicate
formations from appearing in both training and test data. Other useful sensitivity
checks would replace raw frequency by `log1p(frequency)`, remove the most common
nouns a learner probably knows already, and expand the semantic annotations with a
real ontology.

Most importantly, 96% is a benchmark for a much richer statistical model. The
point of this post is to measure how much accuracy can be retained after imposing
a severe human-memory budget.
